Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2023-02-10 manuscript.Rmd
small changes

<!--introduce confirmatory vs. exploratory-->
The scientific community has long pondered the vital distinction between exploration and confirmation, discovery and justification, hypothesis-generating and hypothesis-testing, or prediction and postdiction [@hoyningen-hueneContextDiscoveryContext2006; @shmueliExplainPredict2010, @nosekPreregistrationRevolution2018].
There is a broad consensus that both approaches are necessary for science to progress; exploration for making new discoveries and confirmation for exposing these discoveries to potential falsification and assess in how far it has been corroborated.
Mistaking exploratory findings for empirically confirmed results inflates the the likelihood of believing that there is evidence for a given explanation even if it is false.
A variety of practices, such as researchers' degrees of freedom together with researchers' hindsight bias or naive p-hacking have led to such mistakes becoming commonplace yet unnoticed for a long time.
Recognizing them has led to a crisis of confidence in the empirical sciences [@ioannidisWhyMostPublished2005], and psychology in particular [@opensciencecollaborationEstimatingReproducibilityPsychological2015].
As a response to the crisis, more and more researchers preregister their data collection and analysis plans in advance of the study [@nosekPreregistrationRevolution2018].
They do so to stress the predictive nature of their registered statistical analyses and to produce results that were clearly and transparently to be labeled confirmatory.
Indeed, rigorous application of preregistration prevents researchers from reporting a set of results produced by an arduous process of trial and error as a simple confirmatory story [@wagenmakersAgendaPurelyConfirmatory2012] while keeping low false-positive rates.
This ability to provide a clear distinction between confirmation and exploration has obvious appeal to many who already accepted the practice.
Still, the majority of empirical researchers do not routinely preregister their studies.
One reason may be that some do not find that the theoretical advantages outweigh the practical hurdles, such as specifying every aspect of a theory in advance.
We believe that we can reach a greater acceptance of preregistration by explicating a more general objective of preregistration that benefits all kinds of studies, even those that allow data-dependent decisions.

@@

To arrive at this interpretation, we rely on three arguments.
The first is that theoretical risk is vital for judging evidential support for theories.
The second argument is that the theoretical risk for a given study is generally uncertain.
The third and last argument is that this uncertainty is reduced by applying preregistration.
We conclude that because preregistration decreases uncertainty about the theoretical risk, which in turn increases our expectation to gain evidence for or against a theory, preregistration is potentially useful for any kind of study, no matter how exploratory.

@@

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to theoretical risk, one is bound to overestimate it.
Take, for example, a t-test  of mean differences in two samples.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type-I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item before computing a sum score, broaden, enlarge their definition of the population to be sampled, translate their questionnaires into a different language, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying research hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors influence $P(E|\neg{}H)$.
While, in theory, these factors may leave $P(E|\neg{}H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, they continuously process information about how the study is going, except under strict blinding.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
We, therefore, conclude that statistical properties only guarantee a lower bound to the quantity we seek to minimize.
The only thing we can conclude with some certainty is that theoretical risk is not higher than what the statistical model guarantees without knowledge about the other factors at play.

@@

To illustrate the effect of uncertainty, let $p(E|H) = 0.8$ (e.g., power of 80%) and $p(H) = 0.1$ and assume a uniform distribution for $p(E|\neg{}H)$ of the form:

@@

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, researchers cannot do much to change that.
However, studies do not only differ by how high the theoretical risk is but also how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk, e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed, i.e., alpha error, but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detect ability, prior beliefs, measure of epistemic value) and, therefore, the extend of the benefit very of course with the specifics of the study.
Crucially, even exploratory studies benefit from preregistration, e.g., in this scenario with a $\tau$ of `r some_exploration_tau`, moving from uncertain to certain increases the posterior by a factor of `r round(some_exploration_certain/some_exploration_uncertain, 2)`.

@@

Focusing on uncertainty reduction also explains two common practices of preregistration that do not align with a confirmatory research agenda.
First, researchers seldomly predict precise numerical outcomes, instead they use preregistrations to describe the process that generates the results.
Precise predictions would have very high theoretical risk (they are likely incorrect if the theory is wrong).
A statistical procedure, may have high or low theoretical risk depending on the specifics of the model used.
Specifying the process, therefore, is in line with the rational we propose here, but is incompatible with strictly confirmatory research.
Second, researchers often have to deviate from the preregistration and make data-dependent decisions outside of the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.
However, under our rational some changes may be justified.
Any change would increase the uncertainty about the theoretical risk and may even decrease the theorethical risk.
However, such negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify what to do with missing values where the researchers subsequently encounter missings.
In such case, the detectability is zero, since the data cannot be analyzed without a decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration.
Furthermore, a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory.
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value, likely offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Maybe there is one decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision.

@@

We have proposed a highly specific, yet practical procedure called preregistration as code (PAC) elsewhere [@citepac].
In a PAC, researchers use computer code instead of a verbal description.
This enables a preregistrations to accommodate many different data patterns, yet is very specific about exactly how probable it is to generate evidence even when the theory does not hold (theoretical risk).
To reduce the liklihood of deviation, we suggest that a PAC is tested on simulated or pilot data.
Another advantage is that with a PAC, deviations from the preregistration can be isolated and inspected.
This is possible because in a PAC, preregistration and final manuscript are different versions of the same dynamic document, which can be compared via version control.
However, while PAC is ideally suited to reduce uncertainty about theoretical risk, other more traditional forms of preregistration are also suited to achieve this goal.

>>>>

<!--introduce confirmatory vs. exploratory-->
The scientific community has long pondered the vital distinction between exploration and confirmation, discovery and justification, hypothesis-generating and hypothesis-testing, or prediction and postdiction [@hoyningen-hueneContextDiscoveryContext2006; @shmueliExplainPredict2010, @nosekPreregistrationRevolution2018].
There is a broad consensus that both approaches are necessary for science to progress; exploration for making new discoveries and confirmation for exposing these discoveries to potential falsification and assess in how far it has been corroborated.
Mistaking exploratory findings for empirically confirmed results inflates the the likelihood of believing that there is evidence for a given explanation even if it is false.
A variety of practices, such as researchers' degrees of freedom together with researchers' hindsight bias or naive p-hacking have led to such mistakes becoming commonplace yet unnoticed for a long time.
Recognizing them has led to a crisis of confidence in the empirical sciences [@ioannidisWhyMostPublished2005], and psychology in particular [@opensciencecollaborationEstimatingReproducibilityPsychological2015].
As a response to the crisis, more and more researchers preregister their data collection and analysis plans in advance of the study [@nosekPreregistrationRevolution2018].
They do so to stress the predictive nature of their registered statistical analyses and to produce results that were clearly and transparently to be labeled confirmatory.
Indeed, rigorous application of preregistration prevents researchers from reporting a set of results produced by an arduous process of trial and error as a simple confirmatory story [@wagenmakersAgendaPurelyConfirmatory2012] while keeping low false-positive rates.
This promise of a clear distinction between confirmation and exploration has obvious appeal to many who already accepted the practice.
Still, the majority of empirical researchers do not routinely preregister their studies.
One reason may be that some do not find that the theoretical advantages outweigh the practical hurdles, such as specifying every aspect of a theory in advance.
We believe that we can reach a greater acceptance of preregistration by explicating a more general objective of preregistration that benefits all kinds of studies, even those that allow data-dependent decisions.

@@

To arrive at this interpretation, we rely on three arguments.
The first is that theoretical risk is vital for judging evidential support for theories.
The second argument is that the theoretical risk for a given study is generally uncertain.
The third and last argument is that this uncertainty is reduced by applying preregistration.
We conclude that because preregistration decreases uncertainty about the theoretical risk, which in turn increases our expectation to gain evidence for a theory, preregistration is potentially useful for any kind of study, no matter how exploratory.

@@

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to theoretical risk, one is bound to overestimate it.
Take, for example, a t-test  of mean differences in two samples.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type-I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item before computing a sum score, broaden, enlarge their definition of the population to be sampled, translate their questionnaires into a different language, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying research hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors influence $P(E|\neg{}H)$.
While, in theory, these factors may leave $P(E|\neg{}H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, they continuously process information about how the study is going, except under strict blinding.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
We, therefore, conclude that statistical properties only guarantee a lower bound for theoretical risk.
The only thing we can conclude with some certainty is that theoretical risk is not higher than what the statistical model guarantees without knowledge about the other factors at play.

@@

To illustrate the effect of uncertainty, let $p(E|H) = 0.8$ (e.g., power of 80%) and $p(H) = 0.1$ and assume a uniform distribution for theoretical risk $p(\neg{}E|\neg{}H)$ of the form:

@@

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, researchers cannot do much to change that.
However, studies do not only differ by how high the theoretical risk is but also how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk, e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed, i.e., alpha error, but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of epistemic value) and, therefore, the extend of the benefit very of course with the specifics of the study.
Crucially, even exploratory studies benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau`, so false positive rate of `r 1-some_exploration_tau`, moving from uncertain to certain increases the posterior by a factor of `r round(some_exploration_certain/some_exploration_uncertain, 2)`.

@@

Focusing on uncertainty reduction also explains two common practices of preregistration that do not align with a confirmatory research agenda.
First, researchers seldomly predict precise numerical outcomes, instead they use preregistrations to describe the process that generates the results.
Precise predictions would have very high theoretical risk (they are likely incorrect if the theory is wrong).
A statistical procedure, may have high or low theoretical risk depending on the specifics of the model used.
Specifying the process, therefore, is in line with the rational we propose here, but is incompatible with strictly confirmatory research.
Second, researchers often have to deviate from the preregistration and make data-dependent decisions outside of the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.

However, under our rational some changes may be justified.
Any change would increase the uncertainty about the theoretical risk and may even decrease the theorethical risk.
The changes still may be worthwhile, if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify what to do with missing values where the researchers subsequently encounter missings.
In such case, the detectability is zero, since the data cannot be analyzed without a decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration.
Furthermore, a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory.
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value, likely offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Maybe there is one decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].

@@

We have proposed a highly specific, yet practical preregistration procedure called preregistration as code (PAC) elsewhere [@peikertReproducibleResearchTutorial2021].
In a PAC, researchers use computer code instead of a verbal description.
This enables a preregistrations to accommodate many different data patterns, yet is very specific about exactly how probable it is to generate evidence even when the theory does not hold (theoretical risk).
To reduce the liklihood of deviation, we suggest that a PAC is tested on simulated or pilot data.
Another advantage is that with a PAC, deviations from the preregistration can be isolated and inspected.
This is possible because in a PAC, preregistration and final manuscript are different versions of the same dynamic document, which can be compared via version control.
However, while PAC is ideally suited to reduce uncertainty about theoretical risk, other more traditional forms of preregistration are also suited to achieve this goal.
